---
title: "WV_Thesis_Regression"
author: "Wes Vorley"
date: "2026-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidyquant)
library(minpack.lm)
library(slider)
```

***
#### Read call data, remove N/A values and separate by date call was listed
```{r}
calls <- read.csv("SPX_ATM_calls_MayJul2024.csv") %>%
  na.omit() %>%
  mutate(
    quote_date = as.Date(date), .before = expiration_date,
    expiration_date = as.Date(expiration_date)
  ) %>%
  select(-date)
```

***
#### Obtain daily log returns of SPX from May 2023 to August 2025, add column for 21-trading-day rolling volatility, and annualized using `sqrt(252)`.
```{r}
spx <- tq_get("^GSPC", get = "stock.prices",
               from = "2024-03-01", 
               to = "2024-08-01")  

returns <- spx %>% 
  tq_transmute(select = adjusted, 
               mutate_fun = periodReturn, 
               period = "daily", 
               col_rename = "daily_return") %>%
  mutate(log_return = log1p(daily_return)) %>%
  mutate(
    roll_sd = slide_dbl(log_return, sd, .before = 21, .complete = TRUE),
    mth_vol = roll_sd * sqrt(252)
  ) %>%
  as_tibble()
```

***
#### Function to filter the previous 21 trading days
```{r}
hist_vol <- function(obs_date, lookback = 21) {
  obs_date <- as.Date(obs_date)
  
  returns %>%
    filter(date <= obs_date) %>%
    slice_tail(n = lookback) %>%
    mutate(
      x_day = as.numeric(date - obs_date),
      x = x_day / 252
    )
}
```

***
#### Function to fit linear, quadratic and exponential models to data
```{r}
fit_vol <- function(vol_dat) {
  const <- vol_dat$mth_vol[nrow(vol_dat)]
  
  lin <- lm(mth_vol ~ x, data = vol_dat)
  
  quad <- lm(mth_vol ~ x + I(x^2), data = vol_dat)
  
  # Use log-linear model to get good starting values
  log_fit <- lm(log(mth_vol) ~ x, data = vol_dat)
  start_a <- exp(coef(log_fit)[1])
  start_b <- coef(log_fit)[2]
  exp <- nlsLM(
    mth_vol ~ a * exp(b * x),
    data = vol_dat,
    start = list(a = start_a, b = start_b)
  )
  
  # Return data AND models together in a list
  list(
    vol_dat = vol_dat,
    const = const,
    lin = lin,
    quad = quad,
    exp = exp
  )
}
```

***
#### Functions for integrated variance of each model
```{r}
V_lin <- function(t, fit) {
  # sigma(t) = a*t + b
  a <- coef(fit)[2]   # slope
  b <- coef(fit)[1]   # intercept
  
  res <- (1/3) * a^2 * t^3 + (a * b) * t^2 + b^2 * t
}

V_quad <- function(t, fit) {
  # sigma(t) = a*t^2 + b*t + c
  a <- coef(fit)[3]
  b <- coef(fit)[2]
  c <- coef(fit)[1]
  
  (1/5) * a^2 * t^5 +
    (1/2) * a * b * t^4 +
    ((2/3) * a * c + (1/3) * b^2) * t^3 +
    (b * c) * t^2 +
    c^2 * t
}

V_exp <- function(t, fit) {
  # sigma(t) = a*exp(b*t)
  a <- coef(fit)[1]
  b <- coef(fit)[2]
  
  ifelse(
    abs(b) < 1e-8,
    a^2 * t,
    (a^2 / (2 * b)) * (exp(2 * b * t) - 1)
  )
}
```

***
#### Functions for Black-Scholes model with constant and deterministic volatility
```{r}
bs_const_vol <- function(S, K, r, t, vol) {
  d1 <- (log(S / K) + (r + 0.5 * vol^2) * t) / (vol * sqrt(t))
  d2 <- d1 - vol * sqrt(t)
  
  S * pnorm(d1) - K * exp(-r * t) * pnorm(d2)
}

bs_det_vol <- function(S, K, r, t, V) {
  V <- pmax(V, 1e-10)
  d1 <- (log(S / K) + r * t + 0.5 * V) / sqrt(V)
  d2 <- d1 - sqrt(V)
  
  S * pnorm(d1) - K * exp(-r * t) * pnorm(d2)
}
```

***
#### Function to create a results table for call prices on each date
```{r}
call_tbl_fun <- function(obs_date, vol_dat) {
  call_dat <- calls %>%
    filter(quote_date == obs_date) %>%
    mutate(
      x_day = as.numeric(as.Date(expiration_date) - as.Date(obs_date)),
      x     = x_day / 252,
    ) %>%
    relocate(x_day, .after = expiration_date) %>%
    relocate(x, .after = x_day)
  
  calls_tbl <- tibble(
    x_day       = call_dat$x_day,
    x           = call_dat$x,
    "Obs. Date" = call_dat$quote_date,
    "Exp. Date" = call_dat$expiration_date,
    "S"         = call_dat$spot_price,
    "K"         = call_dat$strike_price,
    "t"         = call_dat$maturity_yrs,
    "r"         = call_dat$interest_rate / 100,
    "True Call" = call_dat$close,
    
    "Const Call" = bs_const_vol(S, K, r, t, vol_dat$const),
    "Lin Call"   = bs_det_vol(S, K, r, t, V_lin(t, vol_dat$lin)),
    "Quad Call"  = bs_det_vol(S, K, r, t, V_quad(t, vol_dat$quad)),
    "Exp Call"   = bs_det_vol(S, K, r, t, V_exp(t, vol_dat$exp)),
    
    # Errors
    err_const = `True Call` - `Const Call`,
    err_lin   = `True Call` - `Lin Call`,
    err_quad  = `True Call` - `Quad Call`,
    err_exp   = `True Call` - `Exp Call`,
    
    abs_err_const = abs(err_const),
    abs_err_lin   = abs(err_lin),
    abs_err_quad  = abs(err_quad),
    abs_err_exp   = abs(err_exp)
  )
  
  return(calls_tbl)
}
```



***
#### Function to create a plot for 21-day rolling volatilities up to a given date
```{r}
plot_vol <- function(obs_date) {
  # Build volatility table
  fit <- hist_vol(obs_date) %>% fit_vol()
  vol_dat <- fit$vol_dat
  
  # Build x-axis breaks
  x_min    <- min(vol_dat$x_day)
  x_max    <- max(vol_dat$x_day)
  x_breaks <- seq(x_min, x_max, length.out = 5)
  x_labels <- format(as.Date(obs_date) + round(x_breaks), "%b %d")
  
  # Get ending date for title
  end_date <- max(as.Date(vol_dat$date))

  ggplot(vol_dat, aes(x = x_day, y = mth_vol)) +
    geom_point(size = 1.5) +
    geom_line(aes(y = fit$const, color = "Constant"), linewidth = 0.8) +
    geom_line(aes(y = predict(fit$lin), color = "Linear"), linewidth = 0.8) +
    geom_line(aes(y = predict(fit$quad), color = "Quadratic"), linewidth = 0.8) +
    geom_line(aes(y = predict(fit$exp), color = "Exponential"), linewidth = 0.8) +
    scale_x_continuous(
      breaks = x_breaks,
      labels = x_labels
    ) +
    labs(
#      title = paste(
#        "SPX 21-Trading-Day Rolling Volatility from",
#        format(as.Date(obs_date), "%B %d"),
#        "to",
#        format(end_date, "%B %d, %Y")
#      ),
      x = "Date",
      y = "Monthly Volatility",
      color = "Model"
    ) +
    theme(plot.title = element_text(hjust = 0.5))
}
```



***
#### Output model performance for volatility fits on all dates
```{r}
vol_errs_tbl <- function(obs_date) {
  fit     <- hist_vol(obs_date) %>% fit_vol()
  vol_dat <- fit$vol_dat

  vol_dat %>%
    transmute(
      `Obs. Date` = as.Date(obs_date),
      date,
      mth_vol,
      fitted_const = fit$const,
      fitted_lin   = predict(fit$lin),
      fitted_quad  = predict(fit$quad),
      fitted_exp   = predict(fit$exp),

      err_const = mth_vol - fitted_const,
      err_lin   = mth_vol - fitted_lin,
      err_quad  = mth_vol - fitted_quad,
      err_exp   = mth_vol - fitted_exp,

      abs_err_const = abs(err_const),
      abs_err_lin   = abs(err_lin),
      abs_err_quad  = abs(err_quad),
      abs_err_exp   = abs(err_exp)
    ) %>%
    select(-c(3:7))
}

vol_errs_tbl("2024-05-27")
vol_errs_tbl("2024-06-03")
vol_errs_tbl("2024-06-10")
vol_errs_tbl("2024-06-17")
vol_errs_tbl("2024-06-24")
vol_errs_tbl("2024-07-01")
vol_errs_tbl("2024-07-08")
vol_errs_tbl("2024-07-15")
vol_errs_tbl("2024-07-22")
vol_errs_tbl("2024-07-29")
```

***
#### Summarize volatility model performance across all dates
```{r}
obs_dates <- c(
  "2024-05-27", "2024-06-03", "2024-06-10", "2024-06-17",
  "2024-06-24", "2024-07-01", "2024-07-08", "2024-07-15",
  "2024-07-22", "2024-07-29"
)

vol_summary_tbl <- map_dfr(obs_dates, function(d) {
  vol_errs_tbl(d) %>%
    summarise(
      Date       = d,
      RMSE_const = sqrt(mean(err_const^2, na.rm = TRUE)),
      RMSE_lin   = sqrt(mean(err_lin^2,   na.rm = TRUE)),
      RMSE_quad  = sqrt(mean(err_quad^2,  na.rm = TRUE)),
      RMSE_exp   = sqrt(mean(err_exp^2,   na.rm = TRUE)),
      MAE_const  = mean(abs_err_const, na.rm = TRUE),
      MAE_lin    = mean(abs_err_lin,   na.rm = TRUE),
      MAE_quad   = mean(abs_err_quad,  na.rm = TRUE),
      MAE_exp    = mean(abs_err_exp,   na.rm = TRUE)
    )
}) %>%
  bind_rows(
    summarise(.,
      Date       = "Mean",
      RMSE_const = round(mean(RMSE_const), 4),
      RMSE_lin   = mean(RMSE_lin),
      RMSE_quad  = mean(RMSE_quad),
      RMSE_exp   = mean(RMSE_exp),
      MAE_const  = mean(MAE_const),
      MAE_lin    = mean(MAE_lin),
      MAE_quad   = mean(MAE_quad),
      MAE_exp    = mean(MAE_exp)
    )
  )

vol_summary_tbl
```

***
#### Best-fit model per date for volatility
```{r}
vol_best_fit_tbl <- vol_summary_tbl %>%
  rowwise() %>%
  mutate(
    RMSE_Best_Fit = c("Constant", "Linear", "Quadratic", "Exponential")[
      which.min(c(RMSE_const, RMSE_lin, RMSE_quad, RMSE_exp))
    ],
    RMSE = min(RMSE_const, RMSE_lin, RMSE_quad, RMSE_exp) %>% round(4),

    MAE_Best_Fit = c("Constant", "Linear", "Quadratic", "Exponential")[
      which.min(c(MAE_const, MAE_lin, MAE_quad, MAE_exp))
    ],
    MAE = min(MAE_const, MAE_lin, MAE_quad, MAE_exp) %>% round(4)
  ) %>%
  ungroup() %>%
  select(Date, RMSE_Best_Fit, RMSE, MAE_Best_Fit, MAE)

vol_best_fit_tbl
```



***
#### Function to create a plot for call prices on a given date
```{r}
plot_call_prices <- function(obs_date) {
  # Build volatility + call price tables
  vol_dat   <- hist_vol(obs_date) %>% fit_vol()
  calls_dat <- call_tbl_fun(obs_date, vol_dat)

  # Build x-axis breaks
  x_min    <- min(calls_dat$x_day)
  x_max    <- max(calls_dat$x_day)
  x_breaks <- seq(x_min, x_max, length.out = 5)
  x_labels <- format(as.Date(obs_date) + round(x_breaks), "%b %d")

  # Get ending date for title
  end_date <- max(as.Date(calls_dat$`Exp. Date`))

  ggplot(calls_dat, aes(x = x_day, y = `True Call`)) +
    geom_point(size = 1.5) +
    geom_line(aes(y = `Const Call`, color = "Constant"), linewidth = 0.8) +
    geom_line(aes(y = `Lin Call`, color = "Linear"), linewidth = 0.8) +
    geom_line(aes(y = `Quad Call`, color = "Quadratic"), linewidth = 0.8) +
    geom_line(aes(y = `Exp Call`, color = "Exponential"), linewidth = 0.8) +
    scale_x_continuous(
      breaks = x_breaks,
      labels = x_labels
    ) +
    labs(
#      title = paste(
#        "Call Prices from",
#        format(as.Date(obs_date), "%B %d"),
#        "to",
#        format(end_date, "%B %d, %Y")
#      ),
      x = "Date",
      y = "Call Price ($)",
      color = "Model"
    ) +
    theme(plot.title = element_text(hjust = 0.5))
}
```

***
#### Plot all volatilities and call prices
```{r}
plot_vol("2024-05-27")
plot_call_prices("2024-05-27")

plot_vol("2024-06-03")
plot_call_prices("2024-06-03")

plot_vol("2024-06-10")
plot_call_prices("2024-06-10")

plot_vol("2024-06-17")
plot_call_prices("2024-06-17")

plot_vol("2024-06-24")
plot_call_prices("2024-06-24")

plot_vol("2024-07-01")
plot_call_prices("2024-07-01")

plot_vol("2024-07-08")
plot_call_prices("2024-07-08")

plot_vol("2024-07-15")
plot_call_prices("2024-07-15")

plot_vol("2024-07-22")
plot_call_prices("2024-07-22")

plot_vol("2024-07-29")
plot_call_prices("2024-07-29")
```

***
#### Output model performance for call prices on all dates
```{r}
errs_tbl <- function(obs_date){
  vol_dat   <- hist_vol(obs_date) %>% fit_vol()
  calls_dat <- call_tbl_fun(obs_date, vol_dat) %>%
    select(-c(1:2), -c(5:13))
  return(calls_dat)
}

errs_tbl("2024-05-27")
errs_tbl("2024-06-03")
errs_tbl("2024-06-10")
errs_tbl("2024-06-17")
errs_tbl("2024-06-24")
errs_tbl("2024-07-01")
errs_tbl("2024-07-08")
errs_tbl("2024-07-15")
errs_tbl("2024-07-22")
errs_tbl("2024-07-29")
```

***
#### Summarize model performance for call prices on all dates
```{r}
summary_tbl <- map_dfr(obs_dates, function(d) {
  errs_tbl(d) %>%
    summarise(
      Date      = d,
      RMSE_const = sqrt(mean(err_const^2, na.rm = TRUE)),
      RMSE_lin   = sqrt(mean(err_lin^2,   na.rm = TRUE)),
      RMSE_quad  = sqrt(mean(err_quad^2,  na.rm = TRUE)),
      RMSE_exp   = sqrt(mean(err_exp^2,   na.rm = TRUE)),
      MAE_const  = mean(abs_err_const, na.rm = TRUE),
      MAE_lin    = mean(abs_err_lin,   na.rm = TRUE),
      MAE_quad   = mean(abs_err_quad,  na.rm = TRUE),
      MAE_exp    = mean(abs_err_exp,   na.rm = TRUE)
    )
}) %>%
  bind_rows(
    summarise(.,
      Date       = "Mean",
      RMSE_const = mean(RMSE_const),
      RMSE_lin   = mean(RMSE_lin),
      RMSE_quad  = mean(RMSE_quad),
      RMSE_exp   = mean(RMSE_exp),
      MAE_const  = mean(MAE_const),
      MAE_lin    = mean(MAE_lin),
      MAE_quad   = mean(MAE_quad),
      MAE_exp    = mean(MAE_exp)
    )
  )

summary_tbl
```

***
#### Best-fit model per date for call prices
```{r}
best_fit_tbl <- summary_tbl %>%
  rowwise() %>%
  mutate(
    RMSE_Best_Fit = c("Constant", "Linear", "Quadratic", "Exponential")[
      which.min(c(RMSE_const, RMSE_lin, RMSE_quad, RMSE_exp))
    ],
    RMSE = min(RMSE_const, RMSE_lin, RMSE_quad, RMSE_exp) %>% round(2),
    
    MAE_Best_Fit = c("Constant", "Linear", "Quadratic", "Exponential")[
      which.min(c(MAE_const, MAE_lin, MAE_quad, MAE_exp))
    ],
    MAE = min(MAE_const, MAE_lin, MAE_quad, MAE_exp) %>% round(2)
  ) %>%
  ungroup() %>%
  select(Date, RMSE_Best_Fit, RMSE, MAE_Best_Fit, MAE)

best_fit_tbl
```

***
#### Paired t-tests: volatility model errors vs. constant baseline
```{r}
# Filter out the "Mean" summary row — keep only the 10 date rows
vol_test_dat <- vol_summary_tbl %>% filter(Date != "Mean")

# RMSE tests (volatility)
vol_rmse_lin  <- t.test(vol_test_dat$RMSE_lin,  vol_test_dat$RMSE_const, paired = TRUE)
vol_rmse_quad <- t.test(vol_test_dat$RMSE_quad, vol_test_dat$RMSE_const, paired = TRUE)
vol_rmse_exp  <- t.test(vol_test_dat$RMSE_exp,  vol_test_dat$RMSE_const, paired = TRUE)

# MAE tests (volatility)
vol_mae_lin   <- t.test(vol_test_dat$MAE_lin,   vol_test_dat$MAE_const,  paired = TRUE)
vol_mae_quad  <- t.test(vol_test_dat$MAE_quad,  vol_test_dat$MAE_const,  paired = TRUE)
vol_mae_exp   <- t.test(vol_test_dat$MAE_exp,   vol_test_dat$MAE_const,  paired = TRUE)

# Compile results into a readable table
vol_ttest_results <- tibble(
  Model    = rep(c("Linear", "Quadratic", "Exponential"), 2),
  Metric   = rep(c("RMSE", "MAE"), each = 3),
  Mean_Diff = c(
    vol_rmse_lin$estimate,  vol_rmse_quad$estimate,  vol_rmse_exp$estimate,
    vol_mae_lin$estimate,   vol_mae_quad$estimate,   vol_mae_exp$estimate
  ),
  t_stat   = c(
    vol_rmse_lin$statistic,  vol_rmse_quad$statistic,  vol_rmse_exp$statistic,
    vol_mae_lin$statistic,   vol_mae_quad$statistic,   vol_mae_exp$statistic
  ),
  p_value  = c(
    vol_rmse_lin$p.value,  vol_rmse_quad$p.value,  vol_rmse_exp$p.value,
    vol_mae_lin$p.value,   vol_mae_quad$p.value,   vol_mae_exp$p.value
  ),
  CI_lower = c(
    vol_rmse_lin$conf.int[1],  vol_rmse_quad$conf.int[1],  vol_rmse_exp$conf.int[1],
    vol_mae_lin$conf.int[1],   vol_mae_quad$conf.int[1],   vol_mae_exp$conf.int[1]
  ),
  CI_upper = c(
    vol_rmse_lin$conf.int[2],  vol_rmse_quad$conf.int[2],  vol_rmse_exp$conf.int[2],
    vol_mae_lin$conf.int[2],   vol_mae_quad$conf.int[2],   vol_mae_exp$conf.int[2]
  )
) %>%
  mutate(across(where(is.numeric), \(x) round(x, 4)))

vol_ttest_results
```

#### Paired t-tests: call pricing model errors vs. constant baseline
```{r}
call_test_dat <- summary_tbl %>% filter(Date != "Mean")

# RMSE tests (call pricing)
call_rmse_lin  <- t.test(call_test_dat$RMSE_lin,  call_test_dat$RMSE_const, paired = TRUE)
call_rmse_quad <- t.test(call_test_dat$RMSE_quad, call_test_dat$RMSE_const, paired = TRUE)
call_rmse_exp  <- t.test(call_test_dat$RMSE_exp,  call_test_dat$RMSE_const, paired = TRUE)

# MAE tests (call pricing)
call_mae_lin   <- t.test(call_test_dat$MAE_lin,   call_test_dat$MAE_const,  paired = TRUE)
call_mae_quad  <- t.test(call_test_dat$MAE_quad,  call_test_dat$MAE_const,  paired = TRUE)
call_mae_exp   <- t.test(call_test_dat$MAE_exp,   call_test_dat$MAE_const,  paired = TRUE)

call_ttest_results <- tibble(
  Model    = rep(c("Linear", "Quadratic", "Exponential"), 2),
  Metric   = rep(c("RMSE", "MAE"), each = 3),
  Mean_Diff = c(
    call_rmse_lin$estimate,  call_rmse_quad$estimate,  call_rmse_exp$estimate,
    call_mae_lin$estimate,   call_mae_quad$estimate,   call_mae_exp$estimate
  ),
  t_stat   = c(
    call_rmse_lin$statistic,  call_rmse_quad$statistic,  call_rmse_exp$statistic,
    call_mae_lin$statistic,   call_mae_quad$statistic,   call_mae_exp$statistic
  ),
  p_value  = c(
    call_rmse_lin$p.value,  call_rmse_quad$p.value,  call_rmse_exp$p.value,
    call_mae_lin$p.value,   call_mae_quad$p.value,   call_mae_exp$p.value
  ),
  CI_lower = c(
    call_rmse_lin$conf.int[1],  call_rmse_quad$conf.int[1],  call_rmse_exp$conf.int[1],
    call_mae_lin$conf.int[1],   call_mae_quad$conf.int[1],   call_mae_exp$conf.int[1]
  ),
  CI_upper = c(
    call_rmse_lin$conf.int[2],  call_rmse_quad$conf.int[2],  call_rmse_exp$conf.int[2],
    call_mae_lin$conf.int[2],   call_mae_quad$conf.int[2],   call_mae_exp$conf.int[2]
  )
) %>%
  mutate(across(where(is.numeric), \(x) round(x, 4)))

call_ttest_results
write.csv(call_ttest_results, "Call_T-Test.csv")
```

